---
title: "Trajectory inferences in all ionocytes"
output: html_document
date: "2024-05-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Load in the packages

```{r }
library(slingshot)
library(Seurat)
library(devtools)
library(cowplot)
library(ggplot2)
library(Matrix)
library(dplyr)
library(RColorBrewer)
library(DelayedMatrixStats)
library(scales)
library(viridis)
library(ComplexHeatmap)
library(circlize)
library(grDevices)
library(PlotTools)
library("fgsea")
library(msigdbr)
```

## load in the Ionocytes (cluster 4 in module 1 analysis) and transform the format that is compatible for slingshot
```{r }

ionocytes = readRDS("ClusteredIono")
DefaultAssay(ionocytes)<-'integrated'
ionocytes.sce <-  as.SingleCellExperiment(ionocytes) 

```

## Trajectory analysis with slingshot
```{r }
sce_slingshot1.ionocytes<- slingshot(ionocytes.sce, clusterLabels = 'leiden_res_0.1', reducedDim = 'TSNE')
SlingshotDataSet(sce_slingshot1.ionocytes) 
summary(sce_slingshot1.ionocytes$slingPseudotime_1)

```
## Plot the pseudotime of ionocytes

```{r }

colors <-  hcl.colors(50, palette = "viridis")
plot(reducedDims(sce_slingshot1.ionocytes)$TSNE, col = colors[cut(sce_slingshot1.ionocytes$slingPseudotime_1,breaks=50)], pch=20,cex=2,asp=1.5)+
lines(SlingshotDataSet(sce_slingshot1.ionocytes),type="lineage",   col = 'black',lwd=2)
title(main = "Tractory inference in ionocytes")
PlotTools::SpectrumLegend(
  "topleft",                             # Legend position
  palette = colors,                     # Display our chosen palette
  legend = seq(60, 0, length.out = 5),  # Annotate positions on legend
  title = "Pseudotime",
  bty = "n"                              # Don't frame with box
)


```


## Test Differential progression
```{r }
ks.test(slingPseudotime(sce_slingshot1.ionocytes)[colData(sce_slingshot1.ionocytes)$type=="CO",1],
        slingPseudotime(sce_slingshot1.ionocytes)[colData(sce_slingshot1.ionocytes)$type=="CF",1]
)
```

## Test trajectory genes
```{r }


pseudo.ionocytes <- TSCAN::testPseudotime(ionocytes.sce, pseudotime=sce_slingshot1.ionocytes$slingPseudotime_1)

pseudo_genes.ionocytes <- subset(pseudo.ionocytes, FDR < 0.05 & abs(logFC)>0.05)

sorted <- pseudo_genes.ionocytes[order(pseudo_genes.ionocytes$FDR,decreasing=FALSE),]


up.ionocytes <- sorted[sorted$logFC > 0,]
down.ionocytes <- sorted[sorted$logFC < 0,]
head(up.ionocytes,10)
head(down.ionocytes,10)

```

## intersect with DEG in part 2
```{r }
DEG_CF_IO=read.csv("ionoCF_CO.csv")
DEG_CF_IO$gene %in% rownames(pseudo_genes.ionocytes)
```

# ionocytes trajectory genes Heatmap visualization
```{r }
## prepare the matrix
DE_gene_iono=rownames(pseudo_genes.ionocytes)
log_count_matrix= ionocytes.sce@assays@data$logcounts
DE_iono_matrix= log_count_matrix[rownames(log_count_matrix) %in% DE_gene_iono,]
### order the cells using their pseudo time
DE_iono_order = slingPseudotime(sce_slingshot1.ionocytes)
DE_iono_order=DE_iono_order[order(DE_iono_order),] ## also has pseudotime info

DE_iono_matrix= DE_iono_matrix[,names(DE_iono_order)]

## prepare the annotation
#### Disease
DE_iono_disease= ionocytes.sce$type
names(DE_iono_disease) = colnames(log_count_matrix)
DE_iono_disease=DE_iono_disease[names(DE_iono_order)]
DE_iono_disease.numerical = as.numeric(ifelse(DE_iono_disease=="CO",0,1))
names(DE_iono_disease.numerical)=names(DE_iono_disease)
#### ionocyte subtype
DE_iono_subtype= ionocytes.sce$leiden_res_0.1
names(DE_iono_subtype) = colnames(log_count_matrix)
DE_iono_subtype=DE_iono_subtype[names(DE_iono_order)]
DE_iono_subtype.numerical = as.numeric(ifelse(DE_iono_subtype=="0",0,1))
names(DE_iono_subtype.numerical)=names(DE_iono_subtype)
```

```{r }

## annotate pseudo time
col_fun = colorRamp2(c(0,30,60), c("darkblue","darkgreen", "yellow"))
ha = HeatmapAnnotation(pseudotime=DE_iono_order,col = list(pseudotime=col_fun))

col_fun_2 = colorRamp2(c(0,1), c("white","black"))
ha_disease = HeatmapAnnotation(CF=DE_iono_disease.numerical,col = list(CF=col_fun_2))
 
col_fun_3 = colorRamp2(c(0,1), c("white","black"))
ha_subtype = HeatmapAnnotation(subtype=DE_iono_subtype.numerical,col = list(subtype=col_fun_3))



Heatmap(as.matrix(DE_iono_matrix), name = "expression",cluster_columns=FALSE,
    bottom_annotation = ha,top_annotation= ha_disease,show_row_names = FALSE,
    show_column_names = FALSE, row_title = NULL, show_row_dend = TRUE)


```

## Gene set enrichment Analysis
```{r }
library("fgsea")

# define the rankings of all the genes
rankings <- sign(pseudo.ionocytes$logFC)*(-log10(pseudo.ionocytes$p.value))
names(rankings) <-rownames(pseudo.ionocytes)

rankings <- sort(rankings, decreasing = TRUE)
min(rankings)
max(rankings)

```

```{r }
## C5 category is according to gene ontology grouping: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4707969/pdf/nihms-743907.pdf
geneSets <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "BP")
### filter background to only include genes that we assessed.
geneSets$gene_symbol <- toupper(geneSets$gene_symbol)
geneSets <- geneSets[geneSets$gene_symbol %in% names(ionocytes.sce),]
m_list <- geneSets %>% split(x = .$gene_symbol, f = .$gs_name)


eaRes <- fgsea(pathways = m_list, stats = rankings, nperm = 5e4, minSize = 10)

ooEA <- order(eaRes$pval, decreasing = FALSE)
head(eaRes[ooEA, 1:3], n = 20)

```

## Plot the geneset enrichment results
```{r }
topPathwaysUp <- eaRes[ES > 0][head(order(pval), n=10), pathway]
topPathwaysDown <- eaRes[ES < 0][head(order(pval), n=10), pathway]
topPathways <- c(topPathwaysUp, rev(topPathwaysDown))
plotGseaTable(m_list[topPathways], rankings, eaRes, 
              gseaParam=0.5)
```

##### collapsed the results, show more statistically significant enriched pathways
```{r }
collapsedPathways <- collapsePathways(eaRes[order(pval)][padj < 0.01], 
                                      m_list, rankings)
mainPathways <- eaRes[pathway %in% collapsedPathways$mainPathways][
                         order(-NES), pathway]
plotGseaTable(m_list[mainPathways], rankings, eaRes, 
              gseaParam = 0.5)


```


